\bf Distortion corrected kernel density estimator on Riemannian manifolds

### 2024-09-10 jcgs/03-chap3-kderm.Rmd
Update paper pdf

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. The is only a small literature dealing with the case where the manifold is unknown. Recently, @berenfeld2021-den, @berenfeld2022-est and @divol2022measure demonstrated the minimax optimality of certain types of KDE on manifolds. In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.

@@

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.

@@

It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in Equation \@ref(eq:twinpeak) distorts the distribution. To recover the true distribution requires the correct Jacobian term for the pushforward from $\bm{v}$ to the volume form of twin peaks manifold. By treating the $\bm{v}$ as an 'output' embedding from input points $\bm{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\bm{\Gamma}_i$ for $i=1,\dots,n$ where $\bm{\Gamma}_i$ is the Riemannian metric of the coordinate system given by $\bm{v}$. This notation is distinct from $\bm{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm. The true density on the manifold can be obtained as $f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}$, where $f(\bm{v}_i)$ is the density of a four-component mixture of normals. Knowledge of $\bm{v}$ and $\bm{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold. \autoref{fig:metadensity} shows the simulated $\bm{v}$ with color indicating the true density of data on the manifold. Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow. The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components. The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\bm{v}$.

@@

\autoref{fig:tpoutliers} summarizes the results. Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively. We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$. Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow. For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions. On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.  The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies. For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes. This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data [@cai2022-tsne].

@@

```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding. The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies. The shading contains all anomalies that are both truly and correctly identified. KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
```

```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding. KDE without distortion correction gives many more misclassified anomalies.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
```

@@

```{r tpisomapvs4ml, fig.pos = "!p", fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row. Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
```

@@

```{r fivedisomapden, fig.align = 'center', out.width = "90%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE. Four manifold learning methods are used row-wise. The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\\% rank region. The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
```

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>%
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>%
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data. DC-KDE has a slightly lower rank correlation for ISOMAP and LLE above 0.96, but significantly outperfoms KDE for Laplacian Eigenmaps and UMAP when more distortion is induced in dimension reduction.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>%
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```

@@

\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE. There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions. The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities. First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings. When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE. As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies. The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times. Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases. Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.

@@

There are several open questions to be explored. The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature. Our method requires the choice of a single scale $r$ that tunes the varying bandwidth matrix. In practice this could be selected using a grid search according to some criteria that depends on the downstream task. For instance for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation. For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition). The second one is that the proposed density estimation method is only limited to the embedding points. Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds [@Berry2017-rz] in the future.
The last question is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way. The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.

>>>>

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. There is only a small literature dealing with the case where the manifold is unknown. Recently, @berenfeld2021-den, @berenfeld2022-est and @divol2022measure demonstrated the minimax optimality of certain types of KDE on manifolds. In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.

@@

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue shades. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

(ref:twinpeaks3dcaption) Scatterplot of the 3-D twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.

@@

It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in Equation \@ref(eq:twinpeak) distorts the distribution. To recover the true distribution requires the correct Jacobian term for the pushforward from $\bm{v}$ to the volume form of twin peaks manifold. By treating the $\bm{v}$ as an 'output' embedding from input points $\bm{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\bm{\Gamma}_i$ for $i=1,\dots,n$ where $\bm{\Gamma}_i$ is the Riemannian metric of the coordinate system given by $\bm{v}$. This notation is distinct from $\bm{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm. The true density on the manifold can be obtained as $f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}$, where $f(\bm{v}_i)$ is the density of a four-component mixture of normals. Knowledge of $\bm{v}$ and $\bm{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold. \autoref{fig:metadensity} shows the simulated $\bm{v}$ with color indicating the true density of data on the manifold. Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in lighter colors. The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components. The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\bm{v}$.

@@

\autoref{fig:tpoutliers} summarizes the results. Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively. We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$. Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors. For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions. On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.  The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies. For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes. This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data [@cai2022-tsne].

@@

```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding of the twin peak data. The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies. The shading contains all anomalies that are both truly and correctly identified. KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
```

```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding of the twin peak data. KDE without distortion correction gives many more misclassified anomalies.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
```

@@

```{r tpisomapvs4ml, fig.pos = "!p", fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities for the twin peak example based on ISOMAP and four other dimension reduction algorithms in each row. Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
# knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
```

@@

```{r fivedisomapden, fig.align = 'center', out.width = "90%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE for the semi-hypersphere example. Four manifold learning methods are used row-wise. The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\\% rank region. The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
```

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>%
  mutate(across(where(is.numeric), ~format(round(., 3), nsmall = 3))) %>%
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data. DC-KDE has a slightly lower rank correlation for ISOMAP and LLE above 0.96, but significantly outperfoms KDE for Laplacian Eigenmaps and UMAP when more distortion is induced in dimension reduction.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>%
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```

@@

\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE. There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions. The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities. First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings. When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE. As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies. The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times. Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases. Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.

@@

There are several open questions to be explored. The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature. Our method requires the choice of a single scale $r$ that tunes the varying bandwidth matrix. In practice, this could be selected using a grid search according to some criteria that depends on the downstream task. For instance, for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation. For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition). The second one is that the proposed density estimation method is only limited to the embedding points. Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds [@Berry2017-rz] in the future.
The last question is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way. The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
